---
title: "STAT 209: Final"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    code_folding: hide
    toc: true
    toc_float: true
  pdf_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
theme_set(theme_bw()) # change theme for ggplot2
knitr::opts_chunk$set(
  tidy    = FALSE,       # display code as typed
  size    = "small",     # slightly smaller font for code
  message = FALSE)  # suppresses some unwanted output
```

# Introduction

You've seen it thousands of times -- someone in a tv show or movie stands on the side of the street with their hands up and somewhat magically a yellow taxi appears to whisk them to their next destination. In a city that is increasingly facing gentrification, industrialization, and has the highest population density of any city in the United States, New York residents must learn to get around the city efficiently. With traffic, expensive parking, and the liability that owning a car in the city presents, I will be using the NYCTaxi database to track taxi usage. I will combine this data with news of the city during the sample time frame, how tourism may affect taxi usage, and if a rise in the number of residents affects the ability to easily travel around the city.

```{r, warning=FALSE}
library(tidyverse)
library(dbplyr)  ## Note the 'b'; this is not dplyr
library(mdsr)    ## Package for our book
library(RMySQL)  ## Standard R/SQL interface package
setwd("~/stat209")
db <- dbConnect_scidb("nyctaxi")
```


```{r}
taxis <- tbl(db, "yellow_old")
```

```{sql connection=db}
SELECT 
  *
  FROM yellow_old
  LIMIT 0, 1;
```


```{sql connection=db}
SELECT 
  str_to_date(pickup_datetime, '%Y-%m-%d') AS pickupdate,
  str_to_date(dropoff_datetime, '%Y-%m-%d') AS dropoffdate,
  pickup_longitude,pickup_latitude,dropoff_longitude,
  dropoff_latitude,fare_amount,tip_amount,vendor_id,passenger_count,total_amount
  FROM yellow_old
  WHERE pickup_latitude > 0 
    AND dropoff_latitude > 0
  GROUP BY pickupdate
  HAVING total_amount > 0;
```
```{r}
query <- 
  "SELECT 
  str_to_date(pickup_datetime, '%Y-%m-%d') AS pickupdate,
  str_to_date(dropoff_datetime, '%Y-%m-%d') AS dropoffdate,
  pickup_longitude,pickup_latitude,dropoff_longitude,
  dropoff_latitude,fare_amount,vendor_id,tip_amount,total_amount,passenger_count,
  SUM(fare_amount) AS total_revenue,
  COUNT(pickup_datetime) AS num_trips
  FROM yellow_old
  WHERE pickup_latitude > 0 
    AND dropoff_latitude > 0
  GROUP BY pickupdate
  HAVING total_amount > 0;"
```

```{r}
delays <- db %>% 
  dbGetQuery(query) %>% 
  collect()

delays
```
#Highest Earning Days
First, I will look at the days in which taxi drivers got the most revenue. The graph below shows that during this month, the total fare amounts were quite variable day to day. The disparity of revenue per day is also quite stark. This data could potentially be used for independent contractors to optimize their time while working -- potentially working a different job during less popular days.

The plot clearly shows that the days of the week that are typically most popular in terms of taxi usage are Thursday, Friday, and Saturday. The most obvious explanation for this is that during these days nightlife is most active in the city. This may increase the number of taxis trips a single person typically takes in a day due to multiple destinations. Another explanation is these days may be popular for tourism.

In conclusion, the most money spent and to be made via taxi fares occur from Thursday-Saturday. Taxi driver can optimize their time by planning their schedules around these days and those using taxis can potentially avoid using taxis during these days as their popularity may increase traffic and surcharges.

```{r}
library(RColorBrewer)
library(lubridate)
```


```{r}
avg = delays$total_revenue

delays <- delays %>%
  mutate(day_of_week = wday(pickupdate, label = TRUE))  %>%
  group_by(day_of_week)

trip_plot <- delays %>%
  ggplot(aes(x = pickupdate, y = total_revenue/1000000, color = day_of_week)) +
  geom_point() +
  geom_hline(yintercept = mean(avg/1000000)) + 
  labs(title = "Highest Earning Days for Taxi Drivers") +
  xlab("Date") +
  ylab("Total Revenue (in millions)") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

trip_plot

p <- delays %>%
  mutate(wday = wday(pickupdate, label = TRUE, week_start = 1)) %>%
  group_by(wday, vendor_id) %>%
  count() %>%
  ggplot(aes(wday, n, colour = vendor_id)) +
  geom_point(size = 4) +
  labs(x = "Day of the week", y = "Total number of pickups") +
  theme(legend.position = "none")
p
```

```{r}
library(dplyr)
library(geosphere)
```

# How Total Distance Affects Trip Numbers

Looking at the days that had the most revenue based on the previous graph. I wanted to see if there was any connection to the length of the trips taken on these days. March 7th and 14th were where the highest earning days while the 3rd and 24th were the least.

The graph is able to account for the impact number of trips may have had on these values. The graph shows that March 7th and 14th both had high levels of both distance and number of trips. March 3rd and 24th however, both had low number of trips. The graph shows that though there may be a positive correlation between total distance driven and revenue, the most significant factor in revenue is the number of trips a driver has in a day. I believe this may be due to the fact that taxis often have a flat fee for driving so smaller distanced trips combined with many trips would increase the total pay for that day. 

```{r}
delays2 <- delays %>%
  rowwise() %>%
  mutate(distance = distHaversine(c(pickup_longitude, pickup_latitude), 
                                  c(dropoff_longitude, dropoff_latitude))) 

distance_plot <- delays2 %>%
  ggplot(aes(x= pickupdate, y = num_trips/1000, size = distance/1000, color = "bcabff")) +
  geom_point() +
  labs(title = "How Total Distance Affects Trip Numbers") +
  xlab("Date") +
  ylab("Number of Trips (in thousands)") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  guides(colour = guide_colourbar(reverse = TRUE))

distance_plot 
```

# How Distance Affects Tip Percentage

I wanted to see if there was a correlation between distance and the amount that a person tips. The data shows that there isn't a strong correlation between these two variables -- but allows us some insight into trends of tipping. It seems that no matter the distance a driver goes, people either tip nothing or tip somewhere in the 20-26% range. Following the past graph, we can conclude that shorter distanced rides again add to the amount of revenue a driver makes in the day when compared to farther distances both due to the potential of flat fees as well as generousity with tips. 
```{r}
tip_percentage <- delays %>%
  rowwise() %>%
  mutate(distance = distHaversine(c(pickup_longitude, pickup_latitude), 
                                  c(dropoff_longitude, dropoff_latitude))) %>%
  mutate(tip_percent = (tip_amount/fare_amount) * 100) %>%
  ggplot(aes(x= tip_percent, y = distance)) +
  geom_point() +
  geom_smooth()

tip_percentage
```
# Discussion
Through this study, I have found that the way for both drivers and riders to be most effective is by paying attention to trends in taxi usage. Drivers tend to be busier during the weekends -- so riders can avoid traffic by using different modes of transportation. Drivers can use this information to see when they are in demand and optimize their time by working more on the weekends and increasing their surcharge. Because shorter trips are often more costly per/minute than longer distances -- it may be more economical for riders to carpool in taxis to their destination. 

Because the dataset is so large -- there are many different angles that further analysis could expand on. Knowing the coordinates of JFK and La Guardia may allow us to see how many people are travelling to the airport, combining this data with one that includes the duration of the trip in time would allow for more information to be told, and plotting more months would allow us to see how these trends change seasonally. 

Though I believe that there is increased taxi usage during the weekend due to nightlife -- this is purely conjecture. More analysis could look to see the most popular times of day that people ride in these taxis. It is also conjecture that many shorter rides is more financial beneficial to drivers due to the fact that the distances in the data were aggregated. Thus -- more analysis is needed to cement this as true.

# Methodology
The first two graphs highlight days of the week by coding them with different colors -- this allows readers to clearly see the trend lines that are present. I decided to add a horizontal line depicting the median in the first graph to show readers the disparity between the lower and upper half of days.

The third graph uses size to code total distance driven to signal to readers clearly how the scale of busier days vary from those slower days. For the first and 3rd graph -- I decided to keep all the dates despite possibly overload of information to showcase more of the data points and potentially clarify the hypotheses. If I were to repeat this, I would remove the years from the labels.

The 4th graph uses a smooth function to show that the line of best fit is sporadic. This highlights that there is little correlation between tip percents and distance driven which draws the eye to the clumps of dots at 0% and 20-26%.
